Deciding Bitvector Termination with Program Synthesis

### 2014-07-03 papers/termination/paper.tex
Conditional loops, nested loops

\noindent {\bf Non-determinism.} A non-termination diagnosis procedure searching for a closed recurrence set under-approximates the transition relation, i.e. 
the closed recurrence set together with the under-approximation characterise non-termination \cite{DBLP:conf/tacas/ChenCFNO14}.
A potential difficulty with such a procedure lies in the handling of non-deterministic assignments. For illustration, consider the following loop:

@@

@@

Sometimes the termination behaviour of a loop depends on its environment.  That is to say,
the loop may not terminate if started in some particular state, but the rest of the program
ensures that such a state cannot be reached on entry to the loop.  The program as a whole
terminates, but if the loop were considered in isolation we would not be able to prove that
it terminates.  We will now consider the problem of encoding the loop's environment as
a synthesis formula.

\noindent {\bf Conditional termination.}
If $L$'s termination depends on the state it begins
executing in, we say that $L$ is \emph{conditionally terminating}. 
The loop we consider now is $L=\langle I,G,T \rangle$; it has an additional predicate $I(x)$ 
representing the states in which the loop can begin executing.
Conditional termination is equivalent to the
formula of Defintion~\ref{def:conditional-termination-formula}. If this formula is satisfiable,
two witnesses are returned:
\begin{itemize}
\item $W$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
guard $G$ is met.
\item $R$ is a ranking function for $L$ as restricted by $W$ -- that is to say, $R$ need only
be well founded on those states satisfying $W \wedge G$.  Since $W$ is an inductive invariant of $L$,
$R$ is strong enough to show that $L$ terminates from any of its initial states.
\end{itemize}

$W$ is called a \emph{supporting invariant} for $L$ and $R$ proves termination relative to $W$.

@@

However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
$R(\langle x,y\rangle ) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
This constitutes a proof that the program as a whole terminates, since the loop always begins
executing in a state that guarantees its termination.

@@

\noindent{\bf Nested loops.}

>>>>

\iffalse
\noindent {\bf Non-determinism.} A non-termination diagnosis procedure searching for a closed recurrence set under-approximates the transition relation, i.e. 
the closed recurrence set together with the under-approximation characterise non-termination \cite{DBLP:conf/tacas/ChenCFNO14}.
A potential difficulty with such a procedure lies in the handling of non-deterministic assignments. For illustration, consider the following loop:

@@

\fi

@@

Sometimes the termination behaviour of a loop depends on its environment.  That is to say,
the loop may not terminate if started in some particular state, but the rest of the program
ensures that such a state cannot be reached on entry to the loop.  The program as a whole
terminates, but if the loop were considered in isolation we would not be able to prove that
it terminates.  We must therefore somehow encode a loop's interaction with the rest of
the program in order to do a sound termination analysis.

Let us assume that we have done some preprocessing of our program which has identified
loops, straight line code blocks and the control flow between these.  In particular,
the control flow analysis has determined which order these code blocks execute in,
and the nesting structure of the loops.

\noindent {\bf Conditional termination.}
If $L$'s termination depends on the state it begins
executing in, we say that $L$ is \emph{conditionally terminating}.
The information we require of the environment is a predicate $I$ which
overapproximates the set of states that $L$ may begin executing in.
That is to say, for each state $x$ that is reachable on entry to $L$,
we have $I(x)$.

Conditional termination is then equivalent to the
formula {\bf [CT]} of Defintion~\ref{def:conditional-termination-formula}. If this formula is satisfiable,
two witnesses are returned:
\begin{itemize}
\item $W$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
guard $G$ is met.
\item $R$ is a ranking function for $L$ as restricted by $W$ -- that is to say, $R$ need only
be well founded on those states satisfying $W \wedge G$.  Since $W$ is an inductive invariant of $L$,
$R$ is strong enough to show that $L$ terminates from any of its initial states.
\end{itemize}

$W$ is called a \emph{supporting invariant} for $L$ and $R$ proves termination relative to $W$.
We require that $I \wedge G$ is strong enough to establish the base case of $W$'s inductiveness.

@@

However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
$R(\langle x,y\rangle) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
This constitutes a proof that the program as a whole terminates, since the loop always begins
executing in a state that guarantees its termination.

\noindent {\bf Nested loops.}
If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
{\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
$L$'s environment, which we model with a relation $S$.  We require that $S$ overapproximates
$L'$ in the sense that it satisfies the following formula:
\begin{align*}
 \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
\end{align*}
Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
the inner loop ($L'$)'s transition relation.

To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
it suffices to summarise the inner loop with the relation:
\begin{align*}
S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
\end{align*}
Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.

\begin{figure}
\begin{lstlisting}
while (x > 0) {
  z = x, y = z + 1;
  while (z > 0) {
    z--, y--;
  }
  x -= y;
}
\end{lstlisting}
\caption{A nested loop\label{fig:nested}}
\end{figure}

\iffalse

@@

\noindent{\bf Nested loops.}
\fi
